Identifying and correcting for misspecifications\\in GWAS summary statistics and polygenic scores

### 2022-06-30 paper/paper-misspec.tex
almost finished with text changes

We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 60\% of the initial time (since only 60\% of the initial non-zero values of the correlation matrix are kept).
Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.9\textwidth]{brca_onco_qc}}
	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$). Only 100,000 HapMap3 variants are represented, at random.}
	\label{fig:brca_onco_qc}
\end{figure}

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.85\textwidth]{res-cad}}
	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics. These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:res_cad}
\end{figure}

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.85\textwidth]{res-vitaminD}}
	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics. These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:res_vitaminD}
\end{figure}

For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}). We believe this is because of the strong regularization used internally. We then run LDpred2-auto-rob for CAD with different shrinkage multiplicative coefficients for the off-diagonal elements of the LD matrix, from 1 (LD matrix unchanged) to 0 (using the identity matrix instead).
When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}). Results are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.

@@

Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
We use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
These were derived using SAIGE \cite[]{zhou2018efficiently}.
First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
We believe this reduction in effective sample size is due to both having related individuals included in the analyses, as well as using SAIGE, as noted in the introduction.
We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
Interestingly, using the small Finnish LD reference panel composed of 503 individuals only we have defined here seems to consistently provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.85\textwidth]{res-FIN}}
	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references. Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:finngen}
\end{figure}

We also use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
Interestingly, when looking at height and to some extent at HDL cholesterol where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}).
Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes  to obtain sufficient sample sizes.

@@

In conclusion, we recommend adopting these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
We also recommend using well-matched LD reference panels. 
More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Methods section \ref{sec:robust}). 
Note that LD blocks are already widely used by several methods, such as lassosum and PRS-CS, because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}.
Imposing independent blocks on the LD matrix could result in further misspecifications \cite[]{zhou2021fast}, but here we have shown that well defined blocks can actually make PGS methods more robust.
PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}.
We believe this is made possible by the use of a strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and the new parameter $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance.
To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
We would like to encourage large biobanks, such as FinnGen and Biobank Japan, to provide LD reference matrices matching the large GWAS summary statistics they provide, ideally based on the same large number of individuals. 
As a future research direction, we are interested in using multi-ancestry-matched LD matrices to use with multi-ancestry GWAS summary statistics in order to improve polygenic prediction in all ancestries.
It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.

>>>>

We form nearly independent LD blocks using the optimal algorithm developed in \cite{prive2021optimal}.
Note that a new parameter \texttt{max\_r2} has been added to this method since, which controls the maximum single squared correlation allowed outside blocks, which offers an extra guarantee that the splitting is good and makes the function much faster by discarding lots of possible splits.
For different numbers of blocks and maximum number of variants in each block, we use the split that minimizes $\big(C2 \cdot \sqrt{5 + C1}\big)$, where $C1$ is the sum of squared correlations outside the blocks, and $C2$ is the sum of squared sizes of the blocks (Figure \ref{fig:cost-split}).
Having a correlation matrix with independent blocks prevents small errors in the algorithm (e.g.\ the Gibbs sampler in LDpred2) from propagating to too many variants.
It also makes running LDpred2 (and lassosum2) faster, taking e.g.\ 70\% of the initial time (since only 70\% of the initial non-zero values of the correlation matrix are kept).
Note that we also compute the correlations within the same blocks and from the same data for PRS-CS.

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.95\textwidth]{brca_onco_qc}}
	\caption{Standard deviations inferred from the OncoArray breast cancer GWAS summary statistics using equation \eqref{eq:approx-sd-log} (\textbf{A:} Raw or \textbf{B:} dividing them by $\sqrt{\text{INFO}}$) versus the ones inferred from the reported GWAS allele frequencies $f_j$ (using $\sqrt{2 f_j (1 - f_j)}$). Only 100,000 HapMap3 variants are represented, at random.}
	\label{fig:brca_onco_qc}
\end{figure}

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.9\textwidth]{res-cad}}
	\caption{Variance explained of CAD in the UK Biobank by PGS derived from external summary statistics. These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:res_cad}
\end{figure}

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.9\textwidth]{res-vitaminD}}
	\caption{Variance explained of vitamin D in the UK Biobank by PGS derived from external summary statistics. These are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation; these values are then squared. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:res_vitaminD}
\end{figure}

For all phenotypes, PRS-CS-auto is very robust, whatever the QC procedure used (e.g.\ see the results for CAD, Figure \ref{fig:res_cad}). We believe this is because of the strong regularization used internally. We then run LDpred2-auto-rob for CAD with different shrinkage multiplicative coefficients for the off-diagonal elements of the LD matrix, from 1 (LD matrix unchanged) to 0 (using the identity matrix instead).
When using a strong shrinkage of 0.3 or 0.4, results for LDpred2-auto-rob are similar whatever the QC used (Figure \ref{fig:res_cad_shrinkage}), as previously observed for PRS-CS-auto. Results with LDpred2-auto-rob are always best when using ``qc2'', the most stringent QC; then no shrinkage (1) seems needed, while a strong shrinkage make predictive performance drop.
For phenotypes with relatively large genetic effects (e.g.\ vitamin D, Figure \ref{fig:res_vitaminD}), PRS-CS-auto perform much worse, possibly due to using too much regularization.

@@

Here we investigate the use of different LD reference panels with GWAS summary statistics from two large biobanks of isolated populations.
We use GWAS summary statistics for five disease endpoints from FinnGen\cite{kurki2022finngen} (release 6), namely breast and prostate cancers (BrCa and PrCa), coronary artery disease (CAD), type 1 and type 2 diabetes (T1D and T2D).
These were derived using SAIGE \cite[]{zhou2018efficiently}.
First, for each phenotype, we estimate a global effective sample using the 80th percentile of imputed sample sizes from equation \eqref{eq:imputeNeff}.
This estimation is only 84.7\% for BrCa, 79.1\% for PrCa, 73.1\% for T1D, 64.5\% for T2D, and 61.3\% for CAD, when compared to the effective sample computed from the reported numbers of cases and controls.
We believe this reduction in effective sample size is due to both having related individuals included in the analyses, as well as using SAIGE, as noted in the introduction.
We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
Interestingly, using the small Finnish LD reference panel we have defined here, composed of 503 individuals only, seems to almost always provide more predictive polygenic scores than using the large UK one composed of 10,000 individuals or the widely used European subset of the 1000 Genomes (1000G) data (Figure \ref{fig:finngen}).
For PRS-CS-auto, no LD reference seems to consistently outperform the two others across all phenotypes considered.
Note that the PGS here are validated in people of UK-like ancestry in the UK biobank to obtain sufficient sample sizes.

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.95\textwidth]{res-FIN}}
	\caption{Results for PGS derived from five FinnGen GWAS summary statistics and using three different LD references. Partial correlations are computed using function \texttt{pcor} of R package bigstatsr where 95\% confidence intervals are obtained through Fisher's Z-transformation, then all values are squared to report the phenotypic variance explained by PGS. Red bars correspond to using the LD with independent blocks (Methods), which is a requirement for PRS-CS.}
	\label{fig:finngen}
\end{figure}

We also use GWAS summary statistics for four continuous outcomes from Biobank Japan \cite[]{sakaue2021cross}.
These were derived using BOLT-LMM-inf \cite[]{loh2018mixed}.
First, for each phenotype, we estimate a global power improvement provided by BOLT-LMM by comparing $\chi^2$-statistics (the boost of BOLT-LMM-inf versus linear regression) across genome-wide significant variants, as recommended in \cite{loh2018mixed}.
This estimated power ratio, which corresponds to the increase in effective sample size, is 1.143 for height, 1.039 for HDL cholesterol, 1.025 for BMI, and 1.013 for systolic blood pressure.
We then compare three different LD reference panels to use with these GWAS summary statistics (Methods section \ref{sec:altLD}).
Interestingly, when looking at height and HDL cholesterol, where we could get the best predictive performance, using the smaller Japanese LD reference we have defined here seems to provide more predictive PGS than using the one using a larger set of East Asian individuals from the UKBB or the widely used East Asian subset of the 1000 Genomes (1000G) data (Figure \ref{fig:bbj}), except when using PRS-CS-auto to predict height.
Note that the PGS here are validated in people of broad East Asian ancestry in the UK biobank and for continuous outcomes to obtain moderate sample sizes for comparison.

\subsection{Recommendations}

Based on previous results from simulations and real-data analyses, we use this section to suggest some recommendations on how to handle misspecifications in GWAS summary statistics and to get the most of polygenic scores. We recall that an overview of these misspecifications, along with possible consequences, and possible corrections is summarized in table \ref{tab:overview}. First, standard quality control should always be performed, e.g.\ variants with low frequency or low imputation quality should be removed; just make sure to recompute MAF and INFO scores for the homogeneous subset of individuals you are interested in. Second, comparison of allele frequencies can be used to e.g.\ detect problems with signs of effects, while comparing standard deviations can be used to detect other issues such as low per-variant GWAS sample sizes.
Third, the LD reference panel should be as close as possible (in terms of genetic ancestry) as the data used to derive the GWAS. In case of any doubt, ancestry proportions can be estimated from GWAS allele frequencies using the method derived in \cite{prive2021using}.
Fourth, if some validation data is available for tuning hyper-parameters, one can use methods such as lassosum2 and LDpred2(-grid). The grid of parameters used in lassosum2 includes a broad range of regularization, which makes it very robust. For LDpred2, we recommend to also include a smaller value for the heritability in the grid of parameters, to allow for more regularization when needed. As these two methods use the same data and the same principle (i.e.\ hyper-parameter tuning), one could run both methods and should the best one when tuning hyper-parameters.
Fifth, if no validation/tuning dataset is available, we recommend using LDpred2-auto with the two new parameters introduced here, or PRS-CS-auto if there are strong misspecifications and the phenotype of interest is known to have small effects only. 
Sixth, we recommend to form independent LD blocks in the 
LD matrix as it proved to make the methods more robust (and slightly faster).

@@

In conclusion, we recommend adopting these changes, i.e.\ performing the (refined) QC proposed here, forming independent LD blocks in the LD matrix, and using more regularization when needed. 
We also recommend using well-matched LD reference panels. 
More regularization can be achieved by testing additional smaller values for the heritability parameter in LDpred2-grid, and using the two new parameters introduced here in LDpred2-auto (Methods section \ref{sec:robust}). 
Note that LD blocks are already widely used by several methods, such as lassosum and PRS-CS, because they allow for processing smaller matrices at once \cite[]{mak2017polygenic,ge2019polygenic}.
Imposing independent blocks on the LD matrix could result in further misspecifications \cite[]{zhou2021fast}, but here we have shown that well defined blocks can actually make PGS methods more robust.
PRS-CS is currently one of the most robust PGS methods; for example, it can use the (small) 1000 Genomes data as LD reference \cite[]{ge2019polygenic}, and can even use a European LD reference panel with multi-ancestry GWAS summary statistics \cite[]{wang2021global}, also shown here.
We believe this is made possible by the use of a strong regularization in PRS-CS ($\phi^{-1} \psi_j^{-1} \ge 1$, which would approximately correspond to using $s=0.5$ in lassosum, $\delta = 1$ in lassosum2, and the new parameter $\texttt{shrink\_corr}=0.5$ in LDpred2-auto).
Using enough regularization is good for robustness, but note that using too much regularization can also damage predictive performance.
To address this limitation, we recommend to rely on a proper QC and choice of ancestry-matched LD instead of on too much regularization.
We would like to encourage large biobanks, such as FinnGen and Biobank Japan, to provide LD reference matrices matching the large GWAS summary statistics they provide, ideally based on the same large number of individuals. 
As a future research direction, we are interested in using multi-ancestry-matched LD matrices to use with multi-ancestry GWAS summary statistics in order to improve polygenic prediction in all ancestries.
It would also be useful to investigate the impact of the QC, well-matched LD reference panels and other adjustments we propose here on other (non-PGS) methods, for which consistency and robustness are likely to be very important as well \cite[]{chen2021improved}.
For example, we are interested in assessing the impact of these misspecifications on the inference of disease architecture parameters in some future work.
